A Plasma Torus Around a Young Low-Mass Star

### 2025-03-18 paper/ms.tex
update

\documentclass{nature3}
\usepackage{graphicx}
\usepackage{float}
\usepackage{verbatim}
\usepackage{hyperref}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{aas_macros_nature}
\usepackage{lineno}

@@

\begin{figure}[!t]
  \centering
  \includegraphics[width=0.7\textwidth]{figures/f1.pdf}
  \caption[]{{\bf Figure 1 (Movie):  TIC~141146667 is a complex periodic
  variable (CPV).}  The online movie,
  available
  \href{https://lgbouma.com/movies/TIC141146667_20250116.mp4}{here},
  shows a baseline of 5{,}784 cycles irregularly sampled over three
  years.  The TESS light curve is phased to the \periodhr\ hour period in
  groups of a few cycles per frame.  This is the period both of
  stellar rotation, and (we hypothesize) of corotating clumps of
  circumstellar material.  Raw data acquired with two minute sampling
  are in gray; black averages to 100 points per cycle.  Similar to
  other stars of this class, the sharp photometric features persist
  for tens to thousands of rotational cycles. }
  \label{fig:lc}
\end{figure}

@@

\begin{figure}[!tp]
  \centering
  \includegraphics[width=0.99\textwidth]{figures/f2.png}
  \caption[]{{\bf Figure 2 (Movie):}  
  Hydrogen emission from circumstellar plasma orbiting TIC 141146667.
  {(\bf TODO)}For the best experience, please view the online movie
  available
  \href{https://lgbouma.com/movies/TIC141146667_sixpanel.mp4}{here}.
  {\bf Panel a:} TESS light curve from UT 2024-02-05 to UT
  2024-02-26 folded on the \periodhr\ hour period.  Black points are
  averaged; gray are the raw data.
  {\bf Panel b:} Keck/HIRES H$\alpha$ spectra
  acquired on UT 2024-02-17.  The continuum is set to unity, and the
  darkest color is set at twice the continuum to accentuate emission
  outside the line core ($|v/v_{\rm eq}|>1$, for $v_{\rm eq}$=130\,\kms).
  While emission in the line core originates in the stellar
  chromosphere, the sinusoidal emission features are most readily
  described by a warped plasma torus.
  {\bf Panel c:} Individual epochs of Panel b, visible in the
  online movie.  The dotted line shows a time-averaged spectrum,
  $f_{\langle t \rangle}$.
  {\bf Panel d:} As in Panel a, but overplotting the
  median-normalized H$\alpha$ light curve at $|v/v_{\rm eq}|<1$.
  {\bf Panel e:} As in Panel b, after subtracting the time-averaged
  spectrum. In addition to circumstellar emission, the line core shows
  absorption during the plasma clump transits.  The asymmetric stretch
  is set to match the dynamic range of the data.
  {\bf Panel f:} Individual epochs of Panel e, visible in the online
  movie.}
  \label{fig:spec}
\end{figure}

@@

Spectra of magetically-active, rapidly rotating stars with a wide range
of masses have been known to exhibit both sinusoidal emission features
\cite{Donati2000,Townsend2005,Dunstone2006,Skelly2008} as well as
transient absorption features in their line cores
\cite{CollierCameron1989,CollierCameron1992,Cang2020} similar to
Figure~\ref{fig:spec}.  No such stars were previously known to show
complex light curves \cite{Bouma2024}.  The usual interpretation for
such spectroscopic variability comes from an analogy to quiescent solar
prominences, which are cool condensations of plasma in the solar corona
that can last days to weeks \cite{VialEngvold2015}.  In our Sun's
magnetosphere, these condensations fall back to the solar surface
because gravity is stronger than any magnetic or centrifugal force
capable of sustaining them.  However for stars with magnetospheric radii
$R_{\rm m}$ that exceed their corotation radii $R_{\rm c}$, the
effective potential experienced by a charged particle can have a local
minimum outside $R_{\rm c}$, enabling the material to be sustained for
much longer timescales \cite{Petit2013,Daley-Yates2024}.  Generally
speaking, such material need neither transit, nor be optically thick.

@@

The other potential analog for the CPVs are the $\sigma$~Ori~E
variables, a rare subset of B stars with radiatively-driven winds that
accumulate into warped plasma tori \cite{Townsend2005,Townsend2008}.
These tori tend to have dense antipodal accumulations of plasma sculpted
by tilted-dipole magnetic fields, and the transits of these clumps are
thought to produce the observed broadband photometric variability
through bound-free scattering \cite{Townsend2005} and Thomson scattering
\cite{Berry2022}.  For $\sigma$~Ori~E and almost all of its analogs, the
result is light curves that appear ``simple'', resembling those of
eclipsing binaries \cite{Townsend2008}.  The two known exceptions,
HD~37776 and HD~64740, show complex light curves resembling CPVs
\cite{Mikulasek2020,Bouma2024} and have spectropolarimetric magnetic
field maps indicating strong contributions from higher-order magnetic
moments \cite{Kochukhov2011,Shultz2018}.  There are two implications:
first, the complexity of CPVs may be a direct consequence of magnetic
fields with highly multipolar contributions.  Second, CPVs could be a
source of astrophysical false positives in photometric searches for
eclipsing binaries and transiting exoplanets around young pre-main
sequence M dwarfs \cite{Johns-Krull2016,Bouma2020}.

@@

\begin{figure}[!b]
  \centering
  \includegraphics[width=0.99\textwidth]{figures/sf1.pdf}
  \caption{Photometric evolution of TIC 141146667 near the Keck/HIRES observation (green bar). 
  {\bf Panel a}: TESS simple aperture photometry. 
  The main data gaps were caused by scattered light from the Earth
  (BTJD 3356-3358.5) and Moon (BTJD 3359.5-3360.5).
  Raw two minute data are in gray; black time-averages ten minute
  sampling.
  {\bf Panels b,c}: Folded light curve before and after spectroscopy.
  Raw two minute data are in gray; black phase-averages to 100 points
  per cycle.
  In the BTJD 3352-3356 window, a state switch occurred near BTJD 3353,
  in which the dip at $\phi$$\approx$0.8 disappeared.
  While the large, complex eclipse feature remained present before and
  after the HIRES sequence, its photometric shape evolved during the
  data gap.
  }
  \label{fig:fulllc}
\end{figure}

>>>>

\documentclass{nature3}
\usepackage{graphicx}
\usepackage{float}
\usepackage{verbatim}
\usepackage{hyperref}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{aas_macros_nature}
\usepackage{lineno}
\usepackage{caption}

@@

\begin{figure}[!t]
  \centering
  \includegraphics[width=0.7\textwidth]{figures/f1.pdf}
  \captionsetup{labelformat=empty}
  \caption{{\bf Figure 1 (Movie):  TIC~141146667 is a complex periodic
  variable (CPV).}  The online movie,
  available
  \href{https://lgbouma.com/movies/TIC141146667_20250116.mp4}{here},
  shows a baseline of 5{,}784 cycles irregularly sampled over three
  years.  The TESS light curve is phased to the \periodhr\ hour period in
  groups of a few cycles per frame.  This is the period both of
  stellar rotation, and (we hypothesize) of corotating clumps of
  circumstellar material.  Raw data acquired with two minute sampling
  are in gray; black averages to 100 points per cycle.  Similar to
  other stars of this class, the sharp photometric features persist
  for tens to thousands of rotational cycles. }
  \label{fig:lc}
\end{figure}

@@

\begin{figure}[!tp]
  \centering
  \includegraphics[width=0.99\textwidth]{figures/f2.png}
  \captionsetup{labelformat=empty}
  \caption{{\bf Figure 2 (Movie):}  
  Hydrogen emission from circumstellar plasma orbiting TIC 141146667.
  {(\bf TODO)}For the best experience, please view the online movie
  available
  \href{https://lgbouma.com/movies/TIC141146667_sixpanel.mp4}{here}.
  {\bf Panel a:} TESS light curve from 5 February 2024 to 26 February
  2024 (UT) folded on the \periodhr\ hour period.  Black points are
  phase-averaged; gray are the raw data.
  {\bf Panel b:} Keck/HIRES H$\alpha$ spectra
  acquired on UT 17 February 2024.  The continuum is set to unity, and the
  darkest color is at twice the continuum to accentuate emission
  outside the line core ($|v/v_{\rm eq}|>1$, for $v_{\rm eq}$=130\,\kms).
  While emission in the line core originates in the stellar
  chromosphere, the sinusoidal emission features are most readily
  described by a warped plasma torus.
  {\bf Panel c:} Individual epochs of Panel b, visible in the
  online movie.  The dotted line shows the time-averaged spectrum,
  $f_{\langle t \rangle}$.
  {\bf Panel d:} As in Panel a, but overplotting the
  median-normalized H$\alpha$ light curve at $|v/v_{\rm eq}|<1$.
  {\bf Panel e:} As in Panel b, after subtracting the time-averaged
  spectrum. In addition to circumstellar emission, the line core shows
  absorption during the plasma clump transits.  The asymmetric stretch
  is set to match the dynamic range of the data.
  {\bf Panel f:} Individual epochs of Panel e, visible in the online
  movie.}
  \label{fig:spec}
\end{figure}

@@

Spectra of magetically-active, rapidly rotating stars with a wide range
of masses have been known to exhibit both sinusoidal emission features
\cite{Donati2000,Townsend2005,Dunstone2006,Skelly2008} as well as
transient absorption features in their line cores
\cite{CollierCameron1989,CollierCameron1992,Cang2020} similar to
Figure~\ref{fig:spec}.  No such stars were previously known to show
complex light curves \cite{Bouma2024}.   One interpretation for
such spectroscopic variability comes from an analogy to quiescent solar
prominences, which are cool condensations of plasma in the solar corona
that can last days to weeks \cite{VialEngvold2015}.  In our Sun's
magnetosphere, these condensations fall back to the solar surface
because gravity is stronger than any magnetic tension or centrifugal force
capable of sustaining them.  However for stars with magnetospheric radii
$R_{\rm m}$ that exceed their corotation radii $R_{\rm c}$, the
effective potential experienced by a charged particle can have a local
minimum outside $R_{\rm c}$, enabling the material to accumulate over
much longer timescales \cite{Petit2013,Daley-Yates2024}.  Generally
speaking, such material need neither transit, nor be optically thick.

@@

The other potential analog for the CPVs are the $\sigma$~Ori~E
variables, a rare subset of B stars with radiatively-driven winds that
accumulate into warped plasma tori \cite{Townsend2005,Townsend2008}.
These tori tend to have dense antipodal accumulations of plasma sculpted
by tilted-dipole magnetic fields, and the transits of these clumps are
thought to produce the observed broadband photometric variability
through bound-free scattering \cite{Townsend2005} and Thomson scattering
\cite{Berry2022}.  For $\sigma$~Ori~E and almost all of its analogs, the
result is light curves that appear ``simple'', resembling those of
eclipsing binaries \cite{Townsend2008}.  The two known exceptions,
HD~37776 and HD~64740, show complex light curves resembling CPVs
\cite{Mikulasek2020,Bouma2024} and have spectropolarimetric magnetic
field maps indicating strong contributions from higher-order magnetic
moments \cite{Kochukhov2011,Shultz2018}.  There are two implications:
first, the complexity of CPVs may be a direct consequence of magnetic
fields with highly multipolar contributions.  Second, CPVs can be a
source of astrophysical false positives in photometric searches for
eclipsing binaries and transiting exoplanets around young pre-main
sequence M dwarfs \cite{Johns-Krull2016,Bouma2020}.

@@

\begin{figure}[!b]
  \centering
  \includegraphics[width=0.99\textwidth]{figures/sf1.pdf}
  \caption{Photometric evolution of TIC 141146667 near the Keck/HIRES observation (green bar). 
  {\bf Panel a}: TESS simple aperture photometry. 
  The main data gaps were caused by scattered light from the Earth
  (BTJD 3356-3358.5) and Moon (BTJD 3359.5-3360.5).
  Raw two minute data are in gray; black time-averages ten minute
  sampling.
  {\bf Panels b,c}: Folded light curve before and after spectroscopy.
  Raw two minute data are in gray; black phase-averages to 100 points
  per cycle.
  During BTJD 3352-3356, a state switch occurred near BTJD 3353,
  in which the dip at $\phi$$\approx$0.8 disappeared.
  While the large eclipse feature remained present before and after
  the HIRES sequence, its photometric shape evolved during the data
  gap.
  }
  \label{fig:fulllc}
\end{figure}
